---
documentclass: jss
author:
  - name: Jason Bryer
    orcid: 0000-0002-2454-0402
    affiliation: City University of New York, School of Professional Studies
    # use this syntax to add text on several lines
    address: |
      | 119 West 31st Street
      | New York, NY 1000
    email: \email{jason.bryer@cuny.edu}
    url: https://bryer.org
  # - name: Second Author
  #   orcid: 0000-0000-0000-0000
  #   affiliation: 'Affiliation \AND'
  #   # To add another line, use \AND at the end of the previous one as above
  # - name: Third Author
  #   orcid: 0000-0000-0000-0000
  #   address: |
  #     | Department of Statistics and Mathematics,
  #     | Faculty of Biosciences,
  #     | Universitat Autònoma de Barcelona
  #   affiliation: |
  #     | Universitat Autònoma 
  #     | de Barcelona
  #   # use a different affiliation in adress field (differently formated here)
  #   affiliation2: Universitat Autònoma de Barcelona
title:
  formatted: "\\pkg{clav}: R Package for Validating Cluster Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "clav: R Package for Validating Cluster Analyses"
  # For running headers, if needed
  short:     "\\pkg{clav}: Cluster Analysis"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
```





# Introduction

This template demonstrates some of the basic LaTeX that you need to know to create a JSS article.

## Code formatting

In general, don't use Markdown, but use the more precise LaTeX commands instead:

* \proglang{Java}
* \pkg{plyr}

One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax).

If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example.

# \proglang{R} code {short-title="R code" #r-code}

Can be inserted in regular R markdown blocks.

```{r}
x <- 1:10
x
```

## Features specific to \pkg{rticles} {short-title="Features specific to rticles"}

* Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
* Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format.
