---
title: "College Readiness Profiles"
author: "Jason Bryer, Ph.D."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{College Readiness Profiles}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  comment = "#>",
  out.width = '100%',
  fig.height = 6,
  fig.width = 9
)
ggplot2::theme_set(ggplot2::theme_minimal())
library(clav)
```


This vignette outlines the analysis for the paper Identifying and Distinguishing College Readiness Profiles Across Academic Outcomes: The Importance of Integrating Academic Skills and Self-Regulated Learning by Timothy J. Cleary, Jason Bryer, Elie ChingYen Yu presented at AERA 2025.

```{r datasetup}
data(daacs)
cluster_vars <- c('Motivation', 'Metacognition', 'Strategies', 'Mathematics', 'Reading', 'Writing')
daacs <- daacs |>
	dplyr::mutate(LogFeedbackViews = log(daacs$FeedbackViews)) |>
	dplyr::mutate(dplyr::across(dplyr::all_of(cluster_vars), clav::scale_this))
```


# Finding the desired number of clusters

```{r optimal-clusters, cache=TRUE}
optimal <- optimal_clusters(daacs[,cluster_vars], max_k = 6)
optimal
```


```{r plot-optimal-clusters}
plot(optimal, ncol = 2)
```

# Validating cluster solution


```{r cluster-validation}
cv <- cluster_validation(daacs[,cluster_vars],
						 n_clusters = 5)
plot(cv, facet = FALSE)
```

```{r plot-distributions}
plot_distributions(cv, plot_in_sample = TRUE, plot_oob_sample = TRUE)
```

```{r bootstrap-validation}
cv_boot <- cluster_validation(daacs[,cluster_vars],
						 n_clusters = 5,
						 sample_size = nrow(daacs),
						 replace = TRUE)
plot(cv_boot, facet = FALSE)
```


# Profile plots

```{r profile-plot}
fit <- stats::kmeans(daacs[,cluster_vars], centers = 5)
profile_plot(daacs[,cluster_vars],
			 clusters = fit$cluster,
			 df_dep = daacs[,c('LogFeedbackViews', 'TermSuccess')],
			 cluster_order = cluster_vars)
```


