---
title: Bootstrapping to Determine the Optimal Number of Clusters
author: Jason Bryer, Ph.D.
output: md_document
self_contained: true
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	error = FALSE,
	fig.height = 4,
	fig.width = 10
)

library(clav)
library(ggplot2)
library(stats)
library(palmerpenguins)

ggplot2::theme_set(ggplot2::theme_minimal())
data(penguins, package = 'palmerpenguins')

cluster_vars <- c('bill_length_mm', 'flipper_length_mm')
penguins <- penguins[complete.cases(penguins[,cluster_vars]),]

penguins <- penguins |>
	dplyr::mutate(dplyr::across(all_of(cluster_vars), clav::scale_this))

```

# Contributions to the field

Cluster analysis is a statistical procedure for grouping observations using an observation-
centered approach as compared to variable-centric approaches (e.g. PCA, factor analysis).
Whether as a preprocessing step for predictive modeling or the primary analysis, validation is
critical for determining generalizability across datasets. The first step for cluster analysis is
determining the optimal number of clusters. Several methods have been proposed to provide
guidance on selecting the correct number of classes (see Table 1), however there is often
disagreement between these methods. This paper proposes a new measure for determining the
optimal number of clusters using bootstrapping. In particular, this approach estimates the
distribution of cluster centers across all variables from boostrap samples. The overlap of cluster
distributions is then calculated. The proposed visualizations provide insight not only into the total
overall but also which cluster and variable combinations may not be well separated by the
clustering model (k means cluster is used here but is generalizable to other methods).

```{r, echo=FALSE}
data(clustering_fit_stats, package = 'clav')
knitr::kable(clustering_fit_stats[,1:3], caption = 'Commonly used metrics for determining the number of clusters.')
```

## Research question

How does bootstrapping compare to classical metrics for determining the optimal number
of clusters?

# Methods

This study will compare a bootstrapping approach to estimating the optimal number of clusters to existing measures (see 1). The bootstrapping method works by taking m bootstrap samples, estimating a k-means model for each bootstrap sample, and recording the mean for each variable and cluster. The result is a collection of distributions for each cluster and variable combination. From those distributions the percent overlap is calculated. The overlap fit metric is the mean or median overlap of all variable and cluster combinations.

## Data source

The Palmer's Penguins [@palmerspenguins] dataset contains observations of penguins on three species (Adelie, Chinstrap, and Gentoo) collected near the Palmer Station in Antarctica. We will attempt to cluster penguins by species using bill and flipper length. Unlike most cluster analysis, the desired classification is known (see Figure \@ref(fig:palmerspenguins-scatter)).

```{r palmerspenguins-scatter, fig.cap='Scatter plot of bill and flipper length by species.', fig.width=10, fig.height=4}
ggplot(penguins, aes(x = bill_length_mm, flipper_length_mm, color = species)) +
	geom_point()
```

# Results

Figure \@ref(fig:fit-statistics) plots six fit statistics described in Table \@ref(tab:fit-statistics-desc) for up to six cluster solutions. There is inconsistency in the desired number of clusters across these methods with only two suggesting a three cluster solution (within group sum of squares and Gap index).

```{r fit-statistics, cache=TRUE, fig.cap='Optimal number cluster plots for six common metrics.', fig.width=10, fig.height=4}
oc <- clav::optimal_clusters(penguins[,cluster_vars], max_k = 6)
plot(oc)
```

Figure \@ref(fig:overlap-fit) summarizes the results of the distribution overlap for each variable and cluster combination for two through six cluster solutions. The green path represents the mean overlap and the red corresponds to the median overlap. All overlap measures are plotted color coded by variable and the black bars correspond to the standard error. Using this method with median, a three cluster solution is appropriate.

```{r, cache=TRUE}
overlap <- cluster_overlap_fit(df =  penguins[,cluster_vars])
```

```{r overlap-fit, fig.cap='Overlap for each cluster and variable combination. Lower overlap is desirable.'}
plot(overlap)
```

To further illustrate how this metric is derived, profile (Figure \@ref(fig:profile-plot)) and distribution (Figure \@ref(fig:distribution-plot)) plots are provided. The profile plot provides the profile structure using both the boostrap sample (left) and predicted class from the out-of-bag sample (right). First, this shows there is stability of the profile structure across multiple samples. Second, we can see that for flipper length there is good separation in the centers across all three clusters, however for bill length there are two clusters that have very similar centers. This can be verified in Figure \@ref(fig:palmerspenguins-scatter). The distribution plot shows the distribution of mean overlap across all boostrap samples. Again, we see that there is good separation across all clusters for flipper length but considerable overlap between two clusters in bill length.

```{r profile-plot, fig.cap='Profile plot plot for k = 3.'}
cv <- cluster_validation(penguins[,cluster_vars], n_clusters = 3)
plot(cv)
```

```{r distribution-plot, fig.cap='Distribution of cluster centers for each variable.'}
plot_distributions(cv)
```

```{r, eval=FALSE}
k1 <- kmeans(penguins[,c('flipper_length_mm'), drop=FALSE], 3)
k2 <- kmeans(penguins[,c('flipper_length_mm', 'bill_length_mm'), drop=FALSE], 3)
table(k1$cluster, penguins$species)
(134 + 91 + 26) / nrow(penguins)
table(k2$cluster, penguins$species)
(146 + 59 + 122) / nrow(penguins)
```

# Practical implications

The overlap fit metric described here provides another tool to use in determining the optimal number of clusters. Unlike the other metrics commonly used, the use of visualizations can provide insight as to whey different cluster solutions may or may not be optimal. And for this example the overlap fit metric would result in a three-cluster solution consistent with the number of species in the original dataset. Future research will explore this method with other datasets where there is a known cluster solution.

# References

Calinski, T., & Harabasz, J. (1974). A dendrite method for cluster analysis. *Communications in Statistics, 3*(1), 1–27. https://doi.org/10.1080/03610927408827101

Carpineto, C., & Romano, G. (2012). Consensus clustering based on a new probabilistic rand index with application to subtopic retrieval. *IEEE Transactions on Pattern Analysis and Machine Intelligence, 34*(12), 2315–2326. https://doi.org/10.1109/TPAMI.2012.80

Davies, D. L., & Bouldin, D. W. (1979). A cluster separation measure. *IEEE Trans. Pattern Anal. Mach. Intell., PAMI-1*(2), 224–227.

Horst, A. M., Hill, A. P., & Gorman, K. B. (2020). *Palmerpenguins: Palmer archipelago (Antarctica) penguin data.* https://doi.org/10.5281/zenodo.3960218

Rousseeuw, P. J. (1987). Silhouettes: A graphical aid to the interpretation and validation of cluster analysis. *J. Comput. Appl. Math., 20*, 53–65.

Theodoridis, S., & Koutroumbas, K. (2008). *Pattern recognition* (4th ed.). USA: Academic Press, Inc.

Thorndike, R. L. (1953). Who belongs in the family? *Psychometrika, 18*(4), 267–276.

Tibshirani, R., Walther, G., & Hastie, T. (2001). Estimating the number of clusters in a data set via the gap statistic. *J. R. Stat. Soc. Series B Stat. Methodol., 63*(2), 411–423.

Ullmann, T., Hennig, C., & Boulesteix, A.-L. (2022). Validation of cluster analysis results on validation data: A systematic framework. *WIREs Data Mining and Knowledge Discovery, 12*(3), e1444. https://doi.org/https://doi.org/10.1002/widm.1444

